---
title: "Joshua French"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: html_notebook
---

# Data exploration (a.k.a., exploratory data analysis)

Based on Chapter 1 of LMWR2, Chapter 1 of ALR4

## Data analysis process

1. Define a question of interest.
2. Collect relevant data.
3. Analyze the data.
4. Interpret your analysis.
5. Make a decision.

"The formulation of a problem is often more essential than its solution, which may be merely a matter of mathematical or experimental skill" - Albert Einstein

### Problem Formulation

* Understand the physical background.
  * Statisticians often work in collaboration with others and need to understand something about the subject area.
* Understand the objective.
  * What are your goals?
  * Make sure you know what the client wants.
* Put the problem into statistical terms.
  * This is often the most challenging step and where irreparable errors are sometimes made.
  * That a statistical method can read in and process the data is not enough. The results of an inapt analysis may be meaningless.

### Data collection

Data collection:
* How the data were collected has a crucial impact on what conclusions can be made.
  * Are the data observational or experimental?
  * Are the data a sample of convenience or were they obtained via a designed sample survey?
* Is there nonresponse bias?
  * The data you do not see may be just as important as the data you do see.
* Are there missing values?
  * This is a common problem that is troublesome and time consuming to handle.
  * How are the data coded? How are the qualitative variables represented?
* What are the units of measurement?
* Beware of data entry errors and other corruption of the data.
  * Perform some data sanity checks.

## Data exploration

An initial exploration of the data should be performed prior to any formal analysis or modeling.

Initial data analysis should consist of numerical summaries and appropriate plots.

### Numerical summaries of data

Statistics can be used to numerically summarize aspects of the data:

* mean
* standard deviation (SD)
* maximum and minimum
* correlation
* other measures, as appropriate

### Visual summaries of data

Plots can provide a useful visual summary of the data.

* For one numerical variable: boxplots, histograms, density plots, etc.
* For two numerical variables: scatterplots.
* For three or more variables, construct interactive and dynamic graphics.
* For one categorial variable: bar charts

Good graphics are essential in data analysis.

* They help us avoid mistakes.
* They help us decide on a model.
* They help communicate the results of our analysis.
* Graphics can be more convincing than text at times.

### What to look for
When summarizing the data, look for:

* outliers
* data-entry errors
* skewness
* unusual distributions
* patterns or structure

## Kidney Example

The National Institute of Diabetes and Digestive and Kidney Diseases conducted a study on 768 adult female Pima Indians living near Phoenix. The following variables were recorded:
* `pregnant` - number of times pregnant
* `glucose` - plasma glucose concentration at 2 hours in an oral glucose tolerance test
* `diastolic` - diastolic blood pressure (mm Hg)
* `triceps` - triceps skin fold thickness (mm)
* `insulin` - 2-hour serum insulin (mu U/ml)
* `bmi` - body mass index (weight in kg/(height in m2))
* `diabetes` - diabetes pedigree function
* `age` - age (years)
* `test` - test whether the patient showed signs of diabetes (coded zero if negative, one if positive).

The data may be obtained from the UCI Repository of machine learning databases at https://archive.ics.uci.edu/ml.

Let's load and examine the structure of the data
```{r}
data(pima, package = "faraway")
str(pima) # structure
head(pima) # first six rows
tail(pima) # last six rows
```
### Numerically summarizing the data
The `summary` command is a useful way to numerically summarize a data frame.

The `summary` function will compute the minimum, 0.25 quantile, mean, median, 0.75 quantile, and maximum of a `numeric` variable.

The `summary` function will count the number of values of each level of a `factor` variable.

Let's summarize the `pima` data frame.

```{r}
summary(pima)
```
### Cleaning the data

Cleaning data involves finding and correcting data quality issues.

Some odd things about the `pima` data

* The minimum blood pressue is zero.
  * That's generally an indication of a health problem.
* The test variable appears to be numerical but should be a categorical variable.
* Many other variables have unusual zeros.
-	Look for anything unusual or unexpected, perhaps indicating a data-entry error.

Let's look at the sorted `diastolic` values.
```{r, echo=T}
sort(pima$diastolic)
```

The first 35 values of `diastolic` are zero. That's a problem.  
* It seems that 0 was used in place of a missing value.
* This is very bad since 0 is a real number and this problem may be overlooked, which can lead to faulty analysis!
* This is why we must check our data carefully for things that don’t make sense.

The value for missing data in R is `NA`.

Several variables share this problem.  Let’s set the 0s that should be missing values to `NA`.

```{r, echo=T}
pima$diastolic[pima$diastolic == 0]  <- NA
pima$glucose[pima$glucose == 0] <- NA
pima$triceps[pima$triceps == 0]  <- NA
pima$insulin[pima$insulin == 0] <- NA
pima$bmi[pima$bmi == 0] <- NA
```

The `test` variable is a categorical variable, not numerical.  

* R thinks the `test` variable is `numeric`.
* In R, a categorical variable is a `factor`.
* We need to convert the `test` variable to a `factor`.

Let's convert `test` to a factor.
```{r, echo=T}
 pima$test <- factor(pima$test)
 summary(pima$test)
```

500 of the cases were negative and 268 were positive.  We can provide more descriptive labels using the `levels` function.

```{r, echo=T}
levels(pima$test) <- c("negative","positive")
```

## Visualizing data with **base** graphics

### Histograms
The `hist` command can be used create a histogram of a numerical vector.

* The labels of the plot can be customized using the `xlab` and `ylab` arguments.
* The main title of the plot can be customized using the `main` argument.
  
  
Here is a slightly customized histogram of diastolic blood pressure.
```{r}
hist(pima$dias, xlab = "diastolic blood pressure" , main="")
```
The histogram is approximately bell-shaped and centered around 70.

We can change the number of breaks in the histogram by specifying the `breaks` argument of the `hist` function.

Consider how the plot changes below.

```{r}
hist(pima$dias, xlab = "diastolic blood pressure", main = "", breaks = 20)
```

### Density plots

Many people prefer the density plot over the histogram because the histogram is more sensitive to its options.  

A density plot is essentially a smoothed version of a histogram.
  * It isn't as blocky.
  * It sometimes has weird things happen at the boundaries.


The `plot` and `density` function can be combined to construct a density plot.
```{r}
plot(density(pima$diastolic, na.rm=TRUE), main = "")
```
In the example above, we have to specify `na.rm = TRUE` so that the density is only computed using the non-missing values.

### Scatter plots

A scatter plot of the sorted numerical values versus their index can be used to identify outliers and see whether the data has many repeated values.

```{r}
plot(sort(pima$diastolic), ylab = "sorted diastolic bp")
```
The flat spots in the plot above show that the `diastolic` variable has mean repeated values.

### Bivariate scatter plots

Bivariate scatter plots can be used to idenify the relationship between two numerical variables.

A scatter plot of diabetes vs diastolic blood pressure.
```{r}
plot(diabetes ~ diastolic, data = pima)
```
There is no clear pattern in the points, so it's difficult to claim a relationship between the two variables.

### Bivariate boxplots

A parallel boxplot of `diabetes` score versus `test` result.
```{r}
plot(diabetes ~ test, data = pima)
```
The median `diabetes` score seems to be a bit higher for positive tests in comparison to the negative tests.

### Multiple plots in one figure

The `par` function can be used to construct multiple plots in one figure. 

* The `mfrow` argument can be used to specify the number of rows and columns of plots you need.

A 1 by 2 set of plots is shown below.
```{r}
par(mfrow = c(1, 2))
plot(diabetes ~ diastolic, data = pima)
plot(diabetes ~ test, data = pima)
```

## Visualizing data with **ggplot2**

The plots we have just created are using the base graphics system in R.
* These are very fast, simple, and professional.

A fancier alternative is to construct plots using the **ggplot2** package.

In its simplest form, **ggplot2** requires you to provide:

* A `ggplot` object that includes the data frame holding the data.
* A mapping arguments that specifies the plot aesthetics (essentially, how the data that will be used in the plot).
* A geometry object that specifies how the aesthetics are used in a plot.

### A **ggplot2** histogram
```{r}
library(ggplot2)
ggpima <-  ggplot(pima)
ggpima + geom_histogram(aes(x=diastolic))
```

### A **ggplot2** density plot
```{r}
ggpima + geom_density(aes(x = diastolic))
```

### A **ggplot2** scatter plot
```{r}
ggpima + geom_point(aes(x = diastolic, y = diabetes))
```

### Customizing **ggplot2** plots 

We can customize our ggplots in numerous ways. Some examples:
* Change the point shapes based on a categorical variable..
* Change the location and orientation of the legend.

```{r}
ggpima + 
  geom_point(aes(x = diastolic, y = diabetes, shape = test)) + 
  theme(legend.position = "top", legend.direction = "horizontal")
```

### Facetting in `ggplot2` 

Facetting creates separate plots (facets) of a data frame based on one or more facetting variables.

Below, we facet the data by the `test` result.

```{r}
ggpima +
  geom_point(aes(x = diastolic, y = diabetes)) +
  facet_grid(~ test)
```

### Summary of **ggplot2**

* Create a `ggplot` object using the `ggplot` function.
  * Specify the data frame the data is contained in (e.g., the data frame is `pima`). 
  * Specify the geometry for the plot (the kind of plot you want to produce)
  * Specify the aesthetics using `aes`.
    * The aesthetic specifies what you see, such as position in the $x$ or $y$ direction or aspects such as shape or color.
    * The aesthetic can be specifed in the geometry, or if you have consistent aesthetics across multiple geometries, in the ``ggplot` statement.
* The advantage of **ggplot2** is more apparent in producing complex plots involving more than two variables.
* A `theme` specifies options for the appearance of the plot.
  * We specified where the `legend` should appear in one plot and to use more than one panel (`facets`) in another. 

## Summary of data exploration

Data exploration helps us to understand the data, identify problems or unusual features of our data, decide on a modeling approach for the data, etc.

You should use both numerical and graphical summaries of data **prior** to modeling the data.