--- 
title: "A progressive introduction to linear models"
author: "Joshua French"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "A collection of R notebooks that progressively introduction how to fit and use linear models."
---
<!-- date: "`r format(Sys.time(), '%Y-%m-%d')`" -->

# Preliminaries {-}

I recommend you execute the following commands install packages we may use in this course.

```{r, eval = FALSE}
# data-related packages
install.packages("palmerpenguins")
# tidyverse
install.packages("tidyverse")
# additional plotting-related packages
install.packages("effects", "plotly")
# manipulation/
# # packages related to books
# books = c("faraway", "alr4", "car", "rms")
# install.packages(books)
# # packages related to tidy/tidying data
# tidy = c("broom", "tidyr", "dplyr", "stringr", "purrr", "tibble", "readr", "forcats")
# install.packages(tidy)
# # packages related to plotting
# moreplots = c("effects", "ggplot2", "ggthemes", "lattice", "HH")
# install.packages(moreplots)
# # packages related to model diagnostics
# diag = c("leaps", "lmtest", "gvlma", "caret")
# install.packages(diag)
# # packages related to workflow
# workflow = c("remotes")
# install.packages(workflow)
```

Lastly, we need to install the **perturb** package, which is currently not available
through the `install.packages` function. To install this from the package developer's
GitHub repository, we run the command below in the Console.

```{r, eval = FALSE}
remotes::install_github(repo = "JohnHendrickx/Perturb")
```

*Acknowledgments*

The **bookdown** package [@R-bookdown] was used to generate this book.

```{r, include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'HistData', 'ggplot2', 'palmerpenguins', 'broom', 'tidyr', 'dplyr', 'stringr', 'purrr', 'tibble', 'readr', 'forcats', 'tidyverse', 'base', 'readxl', 'magrittr', 'ggiraph', 'plotly', 'datasets', 'effects', 'gt'
), 'packages.bib')
```
