---
title: "MATH 4387/5387 Week 2: Data Exploration"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

# Data exploration

Based on Chapter 1 of LMWR2, Chapter 1 of ALR4


```{r, setup, include=FALSE}
# This option means all R code is NOT evaluated
# Delete or comment out if you want to run 
# R commands when knitting
knitr::opts_chunk$set(eval = FALSE)
```

## Data analysis process:

1. Define a statistical question of interest.
2. Collect relevant data.
3. Analyze the data.
4. Interpret your analysis.
5. Make a decision.


### Problem Formulation:

* **Understand the physical background.** Statisticians often work in collaboration with others and need to understand something about the subject area.
* **Understand the objective.** What are your goals? Make sure you know what the collaborator wants.
* **Put the problem into statistical terms.** This is often the most challenging step and where irreparable errors are sometimes made.

### Data collection:

* How the data were collected has a crucial impact on what conclusions can be made. What sampling method? Observational study or randomized controlled experiment?
* Is there nonresponse bias? The data you do not see may be just as important as the data you do see.
* Are there missing values? If so, how (if at all) are these coded?
* What are the variables in the data? What are their units of measurement? Quantitative or qualitative? How are they stored in the data? 
* Beware of data entry errors and other corruption of the data.

## Summaries of Data

An initial exploration of the data should be performed prior to any formal analysis or modeling. Initial data analysis should consist of numerical summaries and appropriate plots.

* Statistics can be used to numerically summarize aspects of the data such as:  mean, median, standard deviation (SD), maximum and minimum, correlation, other measures, as appropriate.

* Plots can provide a useful visual summary of the data.
  * **For one numerical variable:** boxplots, histograms, density plots, etc.
  * **For one categorial variable:** bar charts.
  * **For two numerical variables:** scatter plots.
  * **For one numerical and one categorical variables:** parallel boxplots or density plots that distinguish between category level.
  * **For two categorical variables:** panels of bar charts.
  * **For three or more variables:** one or two variable plots with distinguishing colors or line types, interactive and dynamic graphics.


Good graphics are essential in data analysis.

* They help us to understand our data structure so that we can avoid mistakes.
* They help us decide on a model.
* They help communicate the results of our analysis.
* Depending on the audience, graphics might be the most convincing part of your analysis.

When summarizing the data, look for:

* outliers
* data-entry errors
* skewness
* unusual distributions
* patterns or structure


## Inspecting the Data

```{r}
# Have you already installed the faraway package?  
#install.packages("faraway") # run this line if not.

library(faraway) #loads all datasets
#data(pima, package = "faraway") #loads just one dataset
```

### Understanding Your Data

<font color="red">**Enter a command in the code block below to get help documentation on the `pima` dataset before doing some analysis.** *Note: The eval = FALSE option means this code block will be skipped when you knit and render the entire document to a HTML.*</font>

```{r} 
 # How can you get help?
```


```{r}
str(pima) # structure
head(pima) # first six rows
tail(pima) # last six rows
```

### Numerically Summarize the Data

The `summary` command is a useful way to numerically summarize a data frame.

```{r}
summary(pima)
```

<font color="red">**After looking at the output above, what potential issues do you see with the data?**</font>

### Cleaning the Data

<font color="red">**Explain what the R command below does? Then run it. What potential issues do you see?**</font>

```{r}
sort(pima$diastolic)[1:40]
```


<font color="red">**Replace each ?? in the R code below with the missing value indicator we use in R. Then interpret what each command is doing.**</font>

```{r}
pima$diastolic[pima$diastolic == 0]  <- NA
pima$glucose[pima$glucose == 0] <- NA
pima$triceps[pima$triceps == 0]  <- NA
pima$insulin[pima$insulin == 0] <- NA
pima$bmi[pima$bmi == 0] <- NA
```

<font color="red">**The `test` variable is a categorical variable, not numerical. Explain what the first line of code below is doing.**</font>


```{r}
 pima$test <- factor(pima$test)
 summary(pima$test)
```

<font color="red">**Compare the output from the second line to the original summary we obtained for `test`.**</font>


<font color="blue">We can provide more descriptive labels using the `levels` function.</font>

We change the `0` and `1` levels to `negative` and `positive` to make the data more descriptive. A `summary` of the updates `test` variable shows why this is useful.

```{r}
levels(pima$test) <- c("negative", "positive")
summary(pima$test)
```

## Visualizing data with **`base`** graphics

### Histograms

<font color="red">**The `hist` function can be used create a histogram of a numerical vector. Run the code below and describe the distribution of diastolic values.**</font>

```{r}
hist(pima$diastolic, xlab = "diastolic blood pressure", main = "")
```

<font color="red">**Add an additional option to the `hist` function so the histogram has 20 breaks.**</font>

```{r}
hist(pima$diastolic, ??
     xlab = "diastolic blood pressure", main = "")
```

### Density plots

Many people prefer the **density plot** over the histogram because the histogram is more sensitive to its options.

```{r}
plot(density(pima$diastolic, na.rm = TRUE), main = "",
     type = "l", # "l" is line, "p" is point, and so on
     col = "SeaGreen", #my personal favorite R color
     lwd = 2 # set line width 
     )
```

In the example above, we have to specify `na.rm = TRUE` so that the density is only computed using the non-missing values.

### Index plots

An index plot is a scatter plot of a `numeric` variable versus the index of each value (i.e., the position of the value in the  vector).

* This is most useful for sorted vectors.

```{r}
plot(sort(pima$diastolic), ylab = "sorted diastolic bp")
```

<font color="red">**Give some interesting observations about the diastolic data based on the plot above.**</font>

### Quantile-Quantile Plots

A **Q-Q plot** is a graphical method for comparing two probability distributions by plotting their quantiles against each other. The commands below create a Normal Q-Q plot to help determine whether a normal distribution is a reasonable model.

```{r}
qqnorm(pima$diastolic)
qqline(pima$diastolic)
```

### Bivariate scatter plots

Bivariate scatter plots can be used to identify the relationship between two `numeric` variables.

A **scatter plot** of diabetes vs diastolic blood pressure is shown below.

```{r}
plot(diabetes ~ diastolic, data = pima)
```

<font color="red">**Do you see any pattern in the data? Create another scatter plot to compare another pair of variables in the `pima` dataset.**</font>

### Bivariate boxplots

A parallel boxplot of `diabetes` score versus `test` result is shown below.

```{r}
plot(diabetes ~ test, data = pima)
```

<font color="red">**Do you see any pattern in the data? Create another side-by-side boxplot to compare another pair of variables in the `pima` dataset.**</font>

### Multiple plots in one figure

The `par` function can be used to construct multiple plots in one figure.

* The `mfrow` argument can be used to specify the number of rows and columns of plots you need. 
* As usual in arrays, first number is rows, then columns.

```{r}
par(mfrow = c(1, 2))
plot(diabetes ~ diastolic, data = pima)
plot(diabetes ~ test, data = pima)
par(mfrow = c(1, 1)) # reset to a single plot
```

## Visualizing data with **ggplot2**

The previous plots were created using R's **`base`** graphics system.

* **`base`** graphics are fast and simple to produce while looking professional.

A fancier alternative is to construct plots using the **`ggplot2`** package.

In its simplest form, to construct a (useful) plot in **`ggplot2`**, you need to provide:

* A `ggplot` object.
  * This is usually the object that holds your data frame.
  * The data frame is passed to `ggplot` via the `data` argument.
* A geometry object.
  * Roughly speaking, this is the *kind* of plot you want.
  * e.g., `geom_hist` for a histogram, `geom_point` for a scatter plot, `geom_density` for a density plot.
* An aesthetic mapping.
  * Aesthetic mappings describe how variables in the data are mapped to visual properties of a geometry.
  * This is where you specify which variable with be the `x` variable, the `y` variable, which variable will control color in the plots, etc.
* See https://ggplot2.tidyverse.org for documentation.
* Go to `Help -> Cheat Sheets -> Data Visualization with ggplot2`.

### A **ggplot2** histogram

```{r}
library(ggplot2)
ggpima <-  ggplot(pima)
ggplot(pima) + geom_histogram(aes(x=diastolic))
# ggplot(pima, aes(x = diastolic)) + geom_histogram()
```

* The object is the data ggplot which is loaded first with `ggplot2(pima)`
* The geometry object is specified by `geom_histogram()`
* The aesthetic mapping is plot diastolic blood pressure on the x-axis.

<font color="red">**Use ggplot2 to create a density plot for diastolic blood pressure**</font>

```{r}
ggplot(pima) + geom_density(aes(x=diastolic))
```

<font color="red">**Use ggplot2 to create a scatter plot with diastolic blood pressure on the x axis and diabetes on the y axis.**</font>

```{r}
ggplot(pima)+geom_point(aes(x=diastolic, y=diabetes))
```

### Scaling **ggplot2** plots

In general, *scaling* is the process by which **`ggplot2`** maps variables to unique values. When this is done for discrete variables, **`ggplot2`** will often scale the variable to distinct colors, symbols, or sizes, depending on the aesthetic mapped.

In the example below, we map the `test` variable to the `shape` aesthetic, which is then scaled to different shapes for the different `test` levels.

```{r}
ggpima +
  geom_point(aes(x = diastolic, y = diabetes, shape = test))
```

Alternatively, we can map the `test` variable to the `color` aesthetic, which creates a plot with different colors for each observation based on the `test` level.

```{r}
ggpima +
  geom_point(aes(x = diastolic, y = diabetes, color = test))
```

We can even combine these two aesthetic mappings in a single plot to get different colors and symbols for each level of `test`.

```{r}
ggpima +
  geom_point(aes(x = diastolic, y = diabetes, shape = test, color = test))
```

### Facetting in `ggplot2`

Facetting creates separate panels (facets) of a data frame based on one or more facetting variables.

Below, we facet the data by the `test` result.

```{r}
ggpima +
  geom_point(aes(x = diastolic, y = diabetes)) +
  facet_grid(~ pregnant)
```

### Summary of **ggplot2**

To create a **`ggplot2`** plot:

* Create a`ggplot` object using the `ggplot` function.
  * Specify the data frame the data is contained in (e.g., the data frame is `pima`).
* Specify the geometry for the plot (the kind of plot you want to produce)
* Specify the aesthetics using `aes`.
  * The aesthetic specifies what you see, such as position in the $x$ or $y$ direction or aspects such as shape or color.
  * The aesthetic can be specified in the geometry, or if you have consistent aesthetics across multiple geometries, in the `ggplot` statement.

The advantage of **`ggplot2`** is more apparent in producing complex plots involving more than two variables.

  * **`ggplot2`** makes it easy to plot the data for each group with different colors, symbols, line types, etc.
  * **`ggplot2`** will automatically provide a legend mapping the attributes to the different groups.
  * **`ggplot2`** makes it easy to create separate panels with plots for the observations having a certain characteristic.

## Summary of data exploration

You should use both numerical and graphical summaries of data **prior** to modeling data.

Data exploration helps us to:

* Gain understanding about our data
* Identify problems or unusual features of our data
* Identify patterns in our data
* Decide on a modeling approach for the data
* etc.
