---
title: "Random Vectors and Linear Algebra with R"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r, setup, include=FALSE}
# This option means all R code is NOT evaluated
# Delete or comment out if you want to run 
# R commands when knitting
knitr::opts_chunk$set(eval = FALSE)
```


# Loading the World Bank Data Library

If you have not installed `wbstats` then you will need to uncomment out the first line of code one time.

```{r}
#install.packages("wbstats")
library(wbstats) # World Bank Data
```

# Load most recent literacy rate data
```{r}
edu <- wb_data(
  country = "countries_only",
  indicator = "SE.ADT.LITR.ZS",  # Literacy rate  (% of pop above 15)
  mrnev = 1 #returns most recent non-empty value
)
edu <- edu[,c(3,5)]
names(edu) <- c("country", "literacy_rate")
```


# Load Life Expectancy - SP.DYN.LE00.IN (years)

```{r}
health <- wb_data(
  country = "countries_only", indicator = "SP.DYN.LE00.IN", mrnev = 1)
health <- health[,c(3,5)]
names(health) <- c("country", "life_exp")

# Land Area - AG.LND.TOTL.K2 (sq.km)
area <- wb_data(
  country = "countries_only", indicator = "AG.LND.TOTL.K2", mrnev = 1)
area <- area[,c(3,5)]
names(area) <- c("country", "land_area")
```

# Load GDP per cap (ppp) - NY.GDP.PCAP.PP.CD
```{r}
wealth <- wb_data(
  country = "countries_only",  indicator =  "NY.GDP.PCAP.PP.CD",mrnev = 1)
wealth <- wealth[,c(3,5)]
names(wealth) <- c("country", "gdp_per_cap")
```

# Merge the three separate datasets together and preview

```{r}
country <- merge(edu, merge(health, merge(area, wealth)))
head(country)
```

# Remove country identifiers when we do numerical analysis

```{r}
mydata <- country[,2:5]
```

# Analyzing Correlations Between Pairs

```{r}
#summary(mydata) #summary statistics for each variable alone
var(mydata) #variance matrix
cor(mydata) #correlation matrix
pairs(mydata) #generate pairwise scatterplots
```

```{r}
plot(density(mydata$literacy_rate))
```

# Entering Matrices in R

```{r}
# First Define matrices
a <- matrix(1:6, nrow = 2, ncol = 3, byrow=TRUE)
b <- matrix(c(2, 9, 1, 1, 3, 1), nrow = 2, ncol = 3, byrow=TRUE)
c <- matrix(1:6, nrow = 3, ncol = 2, byrow=FALSE)
```


# Matrices operations in R

```{r}
# First Define matrices
add <- a + b #matrix addition
scalar.prod <- 2*b #matrix times a scalar
diff <- a - scalar.prod #combining operations
matrix.prod <- c %*% b #matrix multiplication uses %*%
trans <- t(c) #transpose of matrix c
```


# Creating special matrices in R

```{r}
d <- diag(c(2, 4, 6, 8), nrow=4, ncol=4) #diagonal matrix
id <- diag(4) #4 by 4 identity matrix
```

```{r}
d.inv <- solve(d) # Find inverse matrix of d
d.det <- det(d) #Compute determinant of matrix d
```