---
pdf_document: default
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  pdf_document: default
title: 'MATH 4387/5387 Week 4: Linear Models'
---

```{r, setup, include=FALSE}
# This option means all R code is NOT evaluated
# Delete or comment out if you want to run 
# R commands when knitting
knitr::opts_chunk$set(eval = FALSE)
```


```{r}    
# install.packages("wbstats")
library(wbstats) # World Bank Data
```

## Load most recent literacy rate

```{r}
edu <- wb_data(
  country = "countries_only",
  indicator = "SE.ADT.LITR.ZS",  # Literacy rate  (% of pop above 15)
  mrnev = 1 #returns most recent non-empty value
)
edu <- edu[,c(3,5)]
names(edu) <- c("country", "literacy_rate")
```

## Life Expectancy - SP.DYN.LE00.IN (years)

```{r}
health <- wb_data(
  country = "countries_only", indicator = "SP.DYN.LE00.IN", mrnev = 1)
health <- health[,c(3,5)]
names(health) <- c("country", "life_exp")
```

## Land Area - AG.LND.TOTL.K2 (sq.km)

```{r}
area <- wb_data(
  country = "countries_only", indicator = "AG.LND.TOTL.K2", mrnev = 1)
area <- area[,c(3,5)]
names(area) <- c("country", "land_area")
```

## GDP per cap (ppp) - NY.GDP.PCAP.PP.CD

```{r}
wealth <- wb_data(
  country = "countries_only",  indicator =  "NY.GDP.PCAP.PP.CD" ,mrnev = 1)
wealth <- wealth[,c(3,5)]
names(wealth) <- c("country", "gdp_per_cap")
```

## Merge the four dataframes together 

```{r}
country <- merge(edu, merge(health, merge(area, wealth)))
head(country)
```

## If you want to compare pairwise correlations

```{r}
summary(country[c(2:5)])
cor(country[c(2:5)])
pairs(country[c(2:5)])
```

## Create dataframe with y first and x second

```{r}
mydata <- data.frame(country$gdp_per_cap, country$literacy_rate)
names(mydata) <- c("gdp_per_cap", "literacy_rate")
```

## Make scatterplot and add regression line

```{r}
plot(gdp_per_cap ~ literacy_rate, data = country,
     pch = 20,
     col=ifelse(literacy_rate > 94.50318 & literacy_rate < 94.5032, "red", "sea green"),
     main = "Scatterplot of Education vs Wealth",
     xlab = "Literacy Rate (%)",
     ylab = "GDP per Capita")
wealth.lm <- lm(mydata)
abline(wealth.lm)
```

## Using a natural log scale

### First we take natural log of gdp_per_cap using `log()`

```{r}
ln.data <- data.frame(log(country$gdp_per_cap), country$literacy_rate)
names(ln.data) <- c("gdp_per_cap", "literacy_rate")
```

### Now do regression using natural log scale

```{r}
ln.lm <- lm(ln.data)

plot(gdp_per_cap ~ literacy_rate, data = ln.data,
     pch = 20,
     col=ifelse(literacy_rate > 94.50318 & literacy_rate < 94.5032, "red", "sea green"),
     main = "Scatterplot of Education vs Wealth",
     xlab = "Literacy Rate (%)",
     ylab = "LN of GDP per Capita")
abline(ln.lm)
```

## Comparing errors using residual plots

```{r}
plot(resid(wealth.lm),
     main = "Plot of Errors Linear Scale",
     xlab = "Index",
     ylab = "Error")

plot(resid(ln.lm),
     main = "Plot of Errors LN Scale",
     xlab = "Index",
     ylab = "Error")
```

```{r}
# For MATH 3191 Slides
train <- country[c(1:3,160),c(2,5)]
lmod <- lm(gdp_per_cap/1000 ~ literacy_rate, data = train)
summary(lmod)
```

```{r}
plot(train$gdp_per_cap/1000 ~ train$literacy_rate, xlab = "Literacy Rate (%)", ylab = "GDP per Capita", xlim=c(0,100), ylim=c(-2, 14))
abline(-1.938, 0.126)
```

