---
title: "Fitting a Linear Model Part 1"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r, setup, include=FALSE}
# This option means all R code is NOT evaluated
# Delete or comment out if you want to run 
# R commands when knitting
knitr::opts_chunk$set(eval = FALSE)
```

# Penguins simple linear regression example

We will use the `penguins` data set in the **palmerpenguins** package [@R-palmerpenguins] to illustrate a very basic simple linear regression analysis. The `penguins` data set was collected and made available by Dr. Kristen Gorman and the Palmer Station, Antarctica LTER, a member of the Long Term Ecological Research Network.

1.  Load the penguins data into R and answer the following questions.

```{r}
#install.packages("palmerpenguins")
library(palmerpenguins)
```

a.  How many observations are in the data?
b.  How many variables? Which variables are categorical and which are numeric?
c.  Do you think the association between bill length of body mass will be positive or negative? Explain why.
d.  Create a scatter plot of where bill length is the response variable and body mass is the predictor.
e.  Do a summary on penguins, and determine whether the data needs to be cleaned.

```{r}
?penguins
nrow(penguins)
ncol(penguins)
```

## Creating a scatter plot

```{r}
plot(bill_length_mm ~ body_mass_g, data = penguins)
```

## Summarizing the Data

We first perform a single linear regression analysis manually using the equations previously provided by regressing `bill_length_mm` on `body_mass_g`.

Using the `summary` function on the `penguins` data frame, we see that both `bill_length_mm` and `body_mass_g` have `NA` values.

```{r}
summary(penguins)
```

## Cleaning the Data

2.  Explain what the code block will do.

```{r}
penguins_clean <- na.omit(penguins[,c("bill_length_mm", "body_mass_g")])
```

## Loading Cleaned Data

We extract the `bill_length_mm` variable from the `penguins` data frame and assign it the name `y` since it will be the response variable. We extract the `body_mass_g` variable from the `penguins` data frame and assign it the name `y` since it will be the predictor variable. We also determine the number of observations and assign that value the name `n`.

```{r}
# extract response and predictor from penguins_clean
y <- penguins_clean$bill_length_mm
x <- penguins_clean$body_mass_g
# determine number of observations
n <- length(y)
```

## Estimating the regression coefficents

We now compute $\hat{\beta}_1$ and $\hat{\beta}_0$. Note that placing `()` around the assignment operations will both perform the assign and print the results.

```{r}
# compute OLS estimates of beta1 and beta0
(b1 <- (sum(x * y) - sum(x) * sum(y) / n)/(sum(x^2) - sum(x)^2/n))
(b0 <- mean(y) - b1 * mean(x))        
```

3.  After running the code above, what are the estimated values for $\widehat{\beta}_0$ and $\widehat{\beta}_1$?
4.  Interpret the practical meaning of these values in terms of penguins.

## Plotting the Line of Best Fit

We can use the `abline` function to overlay the fitted model on the observed data. Note that in simple linear regression, $\hat{\beta}_1$ corresponds to the slope of the fitted line and $\hat{\beta}_0$ will be the intercept.

```{r}
plot(bill_length_mm ~ body_mass_g, data = penguins,
     ylab = "bill length (mm)", xlab = "body mass (g)",
     main = "Penguin size measurements")
# a is the intercept and b is the slope
abline(a = b0, b = b1, col = "red")
```

## Computing the RSS

We can also compute the residuals, $\hat{\epsilon}_1,\ldots,\hat{\epsilon}_n$, the fitted values $\hat{y}_1,\ldots,\hat{y}_n$, and the associated RSS, $RSS=\sum_{i=1}^n \hat{\epsilon}_i^2$.

5. Replace each of the ?? in the R code below to compute the values for $\widehat{y}$,$\widehat{\epsilon}$ and the RSS.

```{r}
yhat <- ?? # compute fitted values
ehat <- ?? # compute residuals
(rss <- ??) # sum of the squared residuals
(sigmasqhat <- rss/(n-2)) # estimated error variance
```

## Comparing with `lm()`

```{r}
## Comparing with `lm()`

(penguin.lm <- lm(penguins_clean)) # calc estimators for b0 and b1
penguin.resid <- resid(penguin.lm) # calc errors
sum(penguin.resid^2) # calc RSS
var(penguin.resid) # compute var of errors
```
