---
title: "Fitting a Linear Model Part 2"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    df_print: paged
---

```{r, setup, include=FALSE}
# This option means all R code is NOT evaluated
# Delete or comment out if you want to run 
# R commands when knitting
knitr::opts_chunk$set(eval = FALSE)
```

# Loading the Galapagos Data

```{r}
#load Galapagos data from faraway package
data(gala, package = "faraway")
str(gala) # basic structure of gala data
```

# Fitting a linear model
Fit a linear model regressing Species on Area, Elevation, Nearest, Scruz, and Adjacent. 

```{r}
# Fit linear model using R.
# Must use data = gala since data vectors are in gala data.frame
lmod <- lm(Species ~ Area + Elevation + Nearest + Scruz + Adjacent, data = gala)
summary(lmod)
```

# Summarizing the Model

```{r}
faraway::sumary(lmod) # the author's modified version of summary
```

```{r}
coef(lmod) # pulls off just the values reg coef.
```

# Construct Matrix Predictors X

```{r}
x <- model.matrix( ~ Area + Elevation + Nearest + Scruz + Adjacent, data = gala)
x
```

# Construct Vector of Observed Response Variable

```{r}
y <- gala$Species
```

# Estimate the regression coefficents

We use the formula

$$\hat{\boldsymbol\beta} = (X^TX)^{-1}(X^T)\mathbf{y}$$
```{r}
# t(x) gives transpose of matrix x
# %*% is matrix multiplication
# solve(a) solves finds inverse matrix of a
solve(t(x) %*% x) %*% t(x) %*% y
```

Which is equivalent to solving the matrix equation
$$ (X^TX)\hat{\boldsymbol\beta} = (X^T)\mathbf{y}$$

```{r}
# solve(a, b) Solves ax = b
solve(t(x) %*% x, t(x) %*% y)
```

Given an $n \times m$ matrix $X$ and an $n \times p$ matrix $Y$, we define the **matrix cross product** as

$$X \times Y = X^TY$$
Using the matrix cross product, we can express the conditions for the regression coefficicents as

$$(X \times X) \hat{\boldsymbol\beta} = X \times \mathbf{y}$$

```{r}
# Even better
solve(crossprod(x), crossprod(x, y))
```


# Quantities we can extract from model object

```{r}
names(lmod)
```

# Estimating the Variance

The OLS estimator for $\mbox{Var} (\boldsymbol\epsilon)$ is given by
$$\hat{\boldsymbol\sigma}^2 = \frac{ \hat{\boldsymbol\epsilon}^T \hat{\boldsymbol\epsilon} }{n-p} = \frac{ \mbox{RSS} }{n-p} .$$
```{r}
e <- residuals(lmod) # Vector of residuals
n <- length(y) # number of observations
p <- 6 # number of predictors

rss <- crossprod(residuals(lmod))
rss/(n-p)
sqrt(rss/(n-p)
```

## Another way to compute RSS

```{r}
deviance(lmod) ## calculates RSS
deviance(lmod)/(n-p) # estimate for variance
```

# Quantities we can extract from summary object

```{r}
lmodsum <- summary(lmod)
names(lmodsum)
```

# Estimating the Variance (Again and Again and ...)


```{r}
sqrt(deviance(lmod)/df.residual(lmod))
lmodsum$sigma
sqrt(sum(residuals(lmod)^2)/df.residual(lmod))
sigma(lmod)
```



# Standard Error of Coefficient Estimates

```{r}
#lmodsum$coef
lmodsum$coef[,2]
```

# Estimated variance/covariance matrix of the regression coefficients

```{r}
vcov(lmod)
```

# Standard Error of Coefficient Estimates (Again)

```{r}
sqrt(diag(vcov(lmod)))
```

# Extracting the fitted values

```{r}
yhat <- fitted(lmod)
```