---
title: "Fitting a Linear Model Part 3"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

# Orthogonality: Oder example

## Loading the Odor Data

Consider the relationship between column temperature, gas/liquid ratio, and packing height in reducing the unpleasant odor of a chemical product that was sold in household use.

```{r}
#load Galapagos data from faraway package
data(odor, package = "faraway")
str(odor) # basic structure of gala data
```

```{r}
head(odor)
```

```{r}
x <- model.matrix(~ temp + gas + pack, 
                  data = odor)
crossprod(x) # matrix of crossproducts
```

## Fitting a linear model
Since the crossproducts of all the column vectors of X are zero, all the regressors are orthogonal to each other and the intercept.


```{r}
lmod <- lm(odor ~ temp + gas + pack, odor)
faraway::sumary(lmod)
```

## Removing an Orthogonal Regressor

Adding or removing `temp` variable doesn’t impact the estimated coefficients.


```{r}
lmod <- lm(odor ~ gas + pack, odor)
faraway::sumary(lmod)
```

# Diabetes Example

403 African Americans were interviewed in a study to understand the prevalence of obesity, diabetes, and other cardiovascular risk factors in central Virginia.  The diabetes data are in the faraway package.
 
We will consider the relationship between cholesterol and several of the predictors. 

## Load the data

```{r}
data(diabetes, package = "faraway")
summary(diabetes)
```

## Fit a Linear Model

```{r}
fm <- chol ~ stab.glu + hdl + glyhb + age + gender + height + weight + waist + hip
lmod <- lm(fm, data = diabetes)
```

```{r}
nrow(diabetes)
```

Here we summarize the results of the model.

```{r}
summary(lmod)
```

## Finding the Fitted Values

```{r}
fitted(lmod) # Shows all 15 fitted values
fitted(lmod)[3] # pulls off on third column
lmod$fitted.values[3] # Another way to do same
```

## Residual plot

```{r}
plot(residuals(lmod) ~ fitted(lmod))
abline(0, 0, lty = 2)
```

## Measuring Response Variation 

We compare $y$ (observed) and $\hat{y}$ (fitted) values.

However, there is a NA for one of the $y$ values in the data, so
note that `lmod$model$chol` are the remaining the $y$ values for the rows without missing data

```{r}
plot(lmod$model$chol ~ fitted(lmod), xlab = "yhat", ylab = "y")
abline(lm(lmod$model$chol ~ fitted(lmod)))
summary(lmod)$r.squared # r-squared
```

# Orthogonality

```{r}
x <- model.matrix(fm, data = diabetes)
crossprod(x)
```