---
title: "Solutions: Interpreting Coefficients Part 2"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r}
#install.packages("alr4") #install package one time
library(alr4) #load datasets from Weisberg text
library(effects)
```

# Example: Fuel Consumption Data Continued

```{r}
data(fuel2001, package = "alr4")
?fuel2001
```

The variables (for the year 2001 unless otherwise noted) are:

- `Drivers`: Number of Licensed drivers in the state
- `FuelC`: Gasoline sold for road use (1000s of gal.)
- `Miles`: Miles of Federal-aid highway miles in the state
- `Pop`: 2001 population age 16 and over
- `Tax`: Gasoline state tax rate (cents/gallon)

## Setting Up a Linear Model

What is the relationship between fuel consumption and various regressors for the 50 United States and the District of Columbia?  

### Adjusting Units

- `Fuel`: Average amount of gasoline sold for road use per person (Gallons/person)
- `Income1K`: Average personal income (in thousands) for the year 2000 per person ($1K/person)
- `Dlic`: Number of licensed drivers per 1000 persons (licensed drivers/1K persons) 

```{r}
# create new regressors/transformed responses to fuel2001 data frame
fuel2001$Fuel <- 1000*fuel2001$FuelC/fuel2001$Pop
fuel2001$Dlic <- 1000*fuel2001$Drivers/fuel2001$Pop
fuel2001$Income1K <- fuel2001$Income/1000
```

```{r}
summary(fuel2001)
```

### Fitting a Model

We will set up a regression model to determine how `Fuel` (gallons per person) is related to `Tax` (cents per gallon),
`Dlic` (drivers per capita), `Income1k` (thousands of dollars of income per capita), and `Miles` (federal highway miles).

$$E( \mbox{Fuel} \ | \ \mbox{Tax, Dlic, Income1K, Miles})=\beta_0+\beta_1 (\mbox{Tax}) +\beta_2 (\mbox{Dlic}) + \beta_3 (\mbox{Income1K}) + \beta_4 \log{(\mbox{Miles})}$$

```{r}
# fit model
lmod <- lm(Fuel ~ Tax + Dlic + Income1K + log(Miles), data = fuel2001)

# summarize model
faraway::sumary(lmod)
```

We see the fitted model is:

$$\widehat{\mbox{E}}(\mbox{Fuel} \ | \ \mbox{Tax, Dlic, Income1k, Miles}) = 154.19 - 4.24 (\mbox{Tax}) + 0.47 (\mbox{Dlic}) - 6.14(\mbox{Income1K}) + 26.76 \log{(\mbox{26.76})}$$

This equation represents the estimated conditional mean of Fuel given fixed values of the regressors Tax, Dlic, Income1K, and Miles.


# Regressors on Logarithmic Scale

Logarithms are commonly used both for the response and for regressors.


### 15. Based on the summary output above, why do you think we used a log scale on `Miles`?

Predictors that span several orders of magnitude should be transformed to the log scale.

In the code block below, we create new variables that are the natural log and log base 10 of Miles and recreate the linear model using each of these new variables.

```{r}
fuel2001$LnMiles <- log(fuel2001$Miles)
fuel2001$LogMiles <- log10(fuel2001$Miles)

lmod.ln <- lm(Fuel ~ Tax + Dlic + Income1K + LnMiles, data = fuel2001)
lmod.log <- lm(Fuel ~ Tax + Dlic + Income1K + LogMiles, data = fuel2001)

faraway::sumary(lmod.ln) #Check that model is the same
faraway::sumary(lmod.log) #Check that model is the same
```

## Below we create an effects plot on a natural log scale.

```{r}
plot(Effect("LnMiles", lmod.ln), 
     main = "ln(Miles) effect plot")
```

## Below we create an effects plot on a log10 scale.

```{r}
plot(Effect("LogMiles", lmod.log), 
     main = "log(Miles) effect plot")
```

## Below we create an effects plot on a regular scale.

```{r}
plot(Effect("Miles", lmod, 
            xlevels = list(Miles = seq(1, 3e5, len = 301))))
```

The effect of increasing Miles is greater in states with fewer miles of roadway, with relatively little change in states with the most roadway. 

This is the usual effect of logarithms: the fitted effect changes most rapidly when the regressor is small and less rapidly when the predictor is large.

## Interpreting Coefficients with Log Scale on Regressor

### Natural Log Scale 

Regressor $X_j$ increasing by 1% while the other regressors remain constant is associated with a $\beta_j/100$ increase in the response variable, on average.

### 16. Interpret the meaning of the coefficient associated to the natural log of Miles which you can find below.

```{r}
summary(lmod.ln)$coefficients[5, 1] #ln coeff
```

We can interpret the regression coefficient for $\log{(\mbox{Miles})}$ as follows. For every 1% increase in the number of miles of Federal highway in a state, the fuel consumption per capita is expected to increase by $0.26755$ gallons per person, on average.

### Common Log (base 10) Scale 

Regressor $X_j$ increasing by a factor of 10 (an increase of 900%) while the other regressors remain constant is associated with a $\beta_j$ increase in the response variable, on average.

### 17. Interpret the meaning of the coefficient associated to the log base 10 of Miles which you can find below.

```{r}
summary(lmod.log)$coefficients[5, 1] #log coeff
```

We can interpret the regression coefficient for $\log_{10}{(\mbox{Miles})}$ as follows. If we increase the number of miles of Federal highway in a state by a factor of 10, the fuel consumption per capita is expected to increase by $61.6$ gallons per person, on average.

# Log-Level Interpretation

It is common for responses to be transformed to a logarithmic scale for theoretical or practical considerations.  

$$\mbox{E}( \log{Y} \ | \ X)= \beta_0 + \beta_1 X.$$

This is sometimes called a **log-level model**. 

- A unit increase in $X$ is associated with a change in the mean $Y$ by the multiplicative effect $\exp^{\beta_1}$.
- **Thus $beta_1$ is the continuous exponential growth/decay rate.**

It is often acceptable to approximate the expected value of a log by the log of the expected value:

$$\log{( \mbox{E}(Y \ | \ X=x) )} \approx\mbox{E}(\log{Y} \ | \ X=x)$$

Thus, we have

$$\mbox{E}(Y \ | \ X=x) \approx e^{\rm{E}(\log{Y} \ | \ X=x)} = e^{\beta_0 + \beta_1 X}=e^{\beta_0}e^{\beta_1X}.$$
# Log-log Interpretation

Consider the log-log simple linear regression model

$$\mbox{E}( \log{Y} \ | \ X) = \beta_0 + \beta_1 \log{X}.$$

When we scale $X$ by a factor of $c$, the response is predicted to grow by a factor of $c^{\beta_1}$, on average.

# Summary of Interpretations (Simple Linear Regression)

**Level-level**: $\mathbf{\mbox{E}(Y \ | \ X=x)} = \boldsymbol\beta_0 + \boldsymbol\beta_1 X$: The predicted change in the response is $\beta_1$ when we increase $X$ by 1 unit, on average.

**Level-log**: $\mathbf{\mbox{E}(Y \ | \ X=x)} = \boldsymbol\beta_0 + \boldsymbol\beta_1 \log{X}$: When we increase $X$ by 1%, the response is predicted to increase by  $\beta_j/100$, on average.

**Log-level**: $\mathbf{\mbox{E}(\log{Y} \ | \ X=x) = \boldsymbol\beta_0 + \boldsymbol\beta_1 X}$: A unit increase in $X$ is predicted to change the response by a factor of $e^{\beta_1}$, on average. The continuous growth rate is $\beta_1$.

**Log-log**: $\mathbf{\mbox{E}(\log{Y} \ | \ X=x) = \boldsymbol\beta_0 + \boldsymbol\beta_1 \log{X}}$: When we scale $X$ by a factor of $c$, the response is predicted to grow by a factor of $c^{\beta_1}$, on average.

# More Practice

1. For a log-level model, interpret the relationship between $X$ and the mean of $Y$ when $X$ increases by 1 unit and $\beta_j=0.3$ and the other predictors do not change.

- The expected value of $Y$ is predicted to increase exponentially with a continuous exponential growth rate of $0.3$ or $30\%$.
- For each increase in $X$ by 1 unit, the expected value of $Y$ is predicted to increase by a factor of $e^{0.3} \approx 1.349859$.
- For each increase in $X$ by 1 unit, the expected value of $Y$ is predicted to grow by $34.99\%$ on average.
- All three of the interpretations above are equivalent.

2. For a log-log model, what is the expected change in $Y$ if we multiply $X$ by a factor of $c$.

- When we scale $X$ by a factor of $c$, the response is predicted to grow by a factor of $c^{\beta_1}$, on average.

3. For a log-log model, interpret the relationship between $X$ and the expected value of $Y$ when $X$ increases by $10\%$ and $\beta_1=0.3$.

- When we scale $X$ by a factor of $1.10$ (a 10% increase), the response is predicted to grow by a factor of $(1.1)^{0.3} \approx 1.029$, on average. 
- Thus, when we increase  $X$ by 10% increase, the response is predicted to grow by by $2.9\%$, on average. 
