---
title: "Categorical Predictors Part 2"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r, setup, include=FALSE}
# This option means all R code is NOT evaluated
# Delete or comment out if you want to run 
# R commands when knitting
knitr::opts_chunk$set(eval = FALSE)
```

```{r, message = FALSE}
# Here we load libraries used in this file
# You may need to install if not already done so
library(alr4) #datasets from Weisberg text
library(ggplot2) #To make some pretty plots
library(lattice) #data visualization
library(HH) #datasets from text by Heiberger and Holland
```


```{r}
levels(UN11$group)
```

```{r}
U1 <- with(UN11, (group == levels(group)[1]) + 0) #oecd =1
U2 <- with(UN11, (group == levels(group)[2]) + 0) #other =1
U3 <- with(UN11, (group == levels(group)[3]) + 0) #africa =1
UN11$U1 = U1
UN11$U2 = U2
UN11$U3 = U3
lm(lifeExpF ~ U2 + U3, data = UN11)
```


# Factors and Quantitative Predictors

It is common to study the effect of a factor AFTER adjusting for one or more other quantitative regressors.  

If we have a mixture of factors and quantitative variables in our data, it’s useful to create a scatterplot of the response versus the covariate that distinguishes the observations for each level.

- This helps us assess whether the relationship between the response and quantitative regressor differs for the levels of the factor.

```{r}
# compute mean lifeExpF for each group
(mean_lifeExpF <- tapply(UN11$lifeExpF, UN11$group, mean))
```

## Exploratory Data Analsis with `ggplot2`

```{r}
ggplot(UN11, aes(x = log(ppgdp), y = lifeExpF, col = group, pch = group)) + geom_point()
```

## Exploratory Data Analsis with `lattice`

```{r}
# A similar plot using the lattice package
library(lattice)
xyplot(lifeExpF ~ log(ppgdp), data = UN11, group = group, auto.key = TRUE)
```

## Exploratory Data Analsis with Base Graphics

```{r}
# using base graphics
# create different colors/symbols for each group
grp_col = as.numeric(UN11$group)

# plot data, using different colors/symbols for each group
plot(lifeExpF ~ log(ppgdp), data = UN11, 
     col = grp_col, pch = grp_col)

# create the legend
legend("topleft", legend = c("oecd", "other", "africa"), col = 1:3, pch = 1:3)
```


### 1. What are some interesting observations/patterns you can see in the plots above?

**Enter answer here**

## Comparing Different Models

Suppose we have a response $Y$, a quantitative regressor $X$, and a two-level factor variable represented by a dummy variable $U$:

$$U = \left\{ \begin{array}{ll} 
0 & \mbox{for the reference level} \\
1 & \mbox{for the treatment level} \end{array} \right.$$

Consider several possible regression models. Note models described extend naturally when the factor has more than 2 levels (as in our case).

- The same one regression line for both levels (combined), $Y= \beta_0 + \beta_1 X + \epsilon$.  
  - In R: `y ~ x`
  - This is a **simple linear model**.

### 2. Enter code to fit a simple linear model with `log(ppgdp)` as the regressor and `lifeExpF` as the response.

```{r}
lmods <- ?? #Enter code here
faraway::sumary(lmods)
```

- A factor predictor but no quantitative predictor, $Y= \beta_0 + \beta_2 U + \epsilon$.  
  - In R: `y ~ u`
  - This is a **one-way model**.

### 3. Enter code to fit a one-way model with `U2` and `U3` as the treatment levels and `lifeExpF` as the response.

```{r}
lmodo <- ?? #Enter code here
faraway::sumary(lmodo)
```

- Separate regression lines for each group having the same slope, $Y= \beta_0+ \beta_1 X+ \beta_2 U + \epsilon$.  
	- In R: `y ~ x + u`.
	- This is known as a **parallel lines** or **main effects model**.
	- $\beta_2$ represents the vertical distance between the regression lines (the effect of the treatment).

### 4. Enter code to fit a main effects model with `log(ppgdp)` as a predictor with `U2` and `U3` as the treatment levels and `lifeExpF` as the response.


```{r}
lmodm <- ?? #Enter code here 
faraway::sumary(lmodm)
```

- Separate lines for each group with different slopes, $Y= \beta_0+ \beta_1 X+ \beta_2 U+ \beta_3 XU + \epsilon$.  
	- In R: `y ~ x + u + x:u` or `y ~ x*u`
	- This is known as a **separate lines** or **interaction model**.
	- `x:u` means the interaction between `x` and `u`.
	- `x*u` means the cross between `x` and `u` (`x`, `u`, and the interaction between `x` and `u`). 

### 5. Enter code to fit an interaction  model with `log(ppgdp)` as a predictor with `U2` and `U3` as the treatment levels and `lifeExpF` as the response

```{r}
lmodi <- ?? #enter code here
faraway::sumary(lmodi)
```

```{r}
coef(lmods)
coef(lmodo)
coef(lmodm)
coef(lmodi)
```


## Model Matrix for the Interaction Model

```{r}
head(model.matrix(lmodi), 10)
```

### 6. How are the last two columns of $X$ computed?

**Enter answer here**

### 7. Find formulas for E(lifeExpF $\vert$ ppgdp = x, group=oecd), E(lifeExpF $\vert$ ppgdp = x, group = other) and E(lifeExpF $\vert$ ppgdp = x, group = africa).

$$\begin{aligned} \widehat{\mbox{lifeExpF}} =& \widehat{\mbox{E}}(\mbox{lifeExpF} \vert \mbox{ppgdp}, \mbox{group})\\
&= \mbox{??} + \mbox{??} \log{(\mbox{ppgdp})} + \mbox{??} U_2 +\mbox{??} U_3 + \mbox{??} \log{(\mbox{ppgdp})}U_2 + \mbox{??} \log{(\mbox{ppgdp})}U_3
\end{aligned}$$

If the group is oecd, then $U_2=0$ and $U_3=0$ giving

$$\begin{aligned}
\widehat{\mbox{lifeExpF}} &= \widehat{\mbox{E}}(\mbox{lifeExpF} \vert \mbox{ppgdp}=x, \mbox{group=oecd})\\
&= \mbox{??} + \mbox{??} \log{(x)}
\end{aligned}$$


If the group is other, then $U_2=1$ and $U_3=0$ giving

$$\begin{aligned}
\widehat{\mbox{lifeExpF}} &= \widehat{\mbox{E}}(\mbox{lifeExpF} \vert \mbox{ppgdp}=x, \mbox{group=other})\\
&= \mbox{??} + \mbox{??} \log{(x)}
\end{aligned}$$

If the group is africa, then $U_2=0$ and $U_3=1$ giving

$$\begin{aligned}
\widehat{\mbox{lifeExpF}} &= \widehat{\mbox{E}}(\mbox{lifeExpF} \vert \mbox{ppgdp}=x, \mbox{group=africa})\\
&= \mbox{??} + \mbox{??} \log{(x)}
\end{aligned}$$

## Plotting the Interaction Model

```{r}
# using ggplot2
ggplot(UN11, aes(x = log(ppgdp), y = lifeExpF, col = group)) + geom_point() + geom_smooth(method = "lm") + theme_bw() + theme(legend.position="top")
```

```{r}
# using lattice
xyplot(lifeExpF ~ log(ppgdp), data = UN11, 
       group = group, type = c("p", "r"), 
       auto.key = TRUE)
```

```{r}
# using base graphics
# create different colors/symbols for each group
grp_col = as.numeric(UN11$group)

# plot data, using different colors/symbols for each group
plot(lifeExpF ~ log(ppgdp), data = UN11, 
     col = grp_col, pch = grp_col)
# fitted lines for each group, 
# changing color for each group.  
# Make sure to match the color number with color number used in plot!
abline(59.21366, 2.24254, col = 1)
abline(59.21366 - 11.17310, 2.24254 + 0.92944, col = 2)
abline(59.21366 - 22.98484, 2.24254 + 1.09498, col = 3)

# create the legend
legend("topleft", legend = c("oecd", "other", "africa"), col = 1:3, pch = 1:3, lty = 1)
```


### 8. Interpret each coefficient in your previous formulas in the context of this dataset.

**Enter answer here**

Note: The interpretation of the **intercept terms** for an interaction model **can be problematic in practice** because the size difference between the coefficients might not represent a typical difference **if the range of the quantitative regressor doesn’t include zero**.

## Interpreting Coefficients in the Interaction Model


**Interpretation is easier if we eliminate the interaction term.** One solution is to center the quantitative regressor at its mean and recompute the model.

- The intercept-related coefficients are interpreted as the effect when the regressor is at its mean value.

## Effects Plot for Interaction Model

```{r}
# Compute effects
ppgdpEffect <- Effect(c("ppgdp", "group"),
                     mod = lmodi,
                     xlevels = list(ppgdp = seq(1, 106000, len = 1000)))
# Plot the effects
# (multiline = TRUE) means all lines plotted on same graph 
plot(ppgdpEffect,
     multiline = TRUE)
```

## Effects Plot for Interaction Model with Log Scale

```{r}
plot(ppgdpEffect,
     rug=FALSE, grid=TRUE, multiline=TRUE,
     xlab = "log(ppgdp)",
     transform.x=list(ppgdp=list(trans=log, inverse=exp)),
     ticks.x =list(ppgdp = list(at= c(100, 1000, 5000, 30000))))
```

```{r}
#Another way to plot on log-scale
plot(ppgdpEffect, multiline=TRUE, grid=TRUE,rug=FALSE,
     axes=list(
       x=list(ppgdp=list(transform=list(trans=log, inverse=exp),
                        lab = "ppdgd, log-scale",
                        ticks=list(at=c(100,1000,5000,30000)),
                        lim = c(1,200000)
                         ))))
```

### 9. What are some interesting observations we can infer from the plot above?

**Enter answer here.**

## Main Effects Model

A **main effects model assumes the slope is the same for all factor levels** but allows each group to have its own intercept.  

```{r}
lmodm <- lm(lifeExpF ~ log(ppgdp) + group, data = UN11)
faraway::sumary(lmodm)
```


### 10. Find formulas for E(lifeExpF $\vert$ ppgdp = x, group=oecd), E(lifeExpF $\vert$ ppgdp = x, group = other) and E(lifeExpF $\vert$ ppgdp = x, group = africa).


$$\begin{aligned} \widehat{\mbox{lifeExpF}} =& \widehat{\mbox{E}}(\mbox{lifeExpF} \vert \mbox{ppgdp}, \mbox{group})\\
&= \mbox{??} + \mbox{??} \log{(\mbox{ppgdp})} + \mbox{??} U_2 +\mbox{??} U_3
\end{aligned}$$

If the group is oecd, then $U_2=0$ and $U_3=0$ giving

$$\begin{aligned}
\widehat{\mbox{lifeExpF}} &= \widehat{\mbox{E}}(\mbox{lifeExpF} \vert \mbox{ppgdp}=x, \mbox{group=oecd})\\
&= \mbox{??} + \mbox{??} \log{(x)}
\end{aligned}$$


If the group is other, then $U_2=1$ and $U_3=0$ giving

$$\begin{aligned}
\widehat{\mbox{lifeExpF}} &= \widehat{\mbox{E}}(\mbox{lifeExpF} \vert \mbox{ppgdp}=x, \mbox{group=other})\\
&= \mbox{??} + \mbox{??} \log{(x)}
\end{aligned}$$

If the group is africa, then $U_2=0$ and $U_3=1$ giving

$$\begin{aligned}
\widehat{\mbox{lifeExpF}} &= \widehat{\mbox{E}}(\mbox{lifeExpF} \vert \mbox{ppgdp}=x, \mbox{group=africa})\\
&= \mbox{??} + \mbox{??} \log{(x)}
\end{aligned}$$

- Main effects models are easier to interpret since the effect of the quantitative regressor is the same for all levels of the factor.
- This model is called the Analysis of Covariance (ANCOVA) when the factor levels indicate a randomly assigned treatment for subjects.


## Plotting the Main Effects Model

```{r}
# using ggplot2

# this limits the line to the range of the observed data (avoiding extrapolation)
predx <- unlist(tapply(log(UN11$ppgdp), UN11$group, function(x) seq(min(x), max(x), len = 2)))

# associated group for each x-value
g = rep(c("oecd", "other", "africa"), each = 2)

# predict response for each combination of x, group
predy = predict(lmodm, newdata = data.frame(ppgdp = exp(predx), group = g))

# combine together in new data frame
fitted_df = data.frame(ppgdp = exp(predx), group = g, lifeExpF = predy)

# plot main effects/ancova model using ggplot2
ggplot(UN11, aes(x = log(ppgdp), y = lifeExpF, col = group)) + geom_point() + geom_line(data = fitted_df, aes(x = log(ppgdp), y = lifeExpF)) + theme_bw()  + theme(legend.position="top")
```

```{r}
# using lattice package (essentially)
#library(HH) #you will need to install this package
# doesn't like log(ppgdp), so we need to create this variable
UN11$lppgdp = log(UN11$ppgdp)

ancovaplot(lifeExpF ~ lppgdp + group, data = UN11)
```

```{r}
# using base graphics
# create different colors/symbols for each group
grp_pch = as.numeric(UN11$group)
grp_col = grp_pch + 3

# plot data, using different colors/symbols for each group
plot(lifeExpF ~ log(ppgdp), data = UN11, col = grp_col, pch = grp_pch)

# fitted lines for each group, changing color for each group.  
# Be sure to match the color number with color number used in plot!
abline(49.53, 3.2, col = 4)
abline(48, 3.2, col = 5)
abline(37.36, 3.2, col = 6)

# create the legend
legend("topleft", legend = c("oecd", "other", "africa"),
       col = 4:6, pch = 1:3, lty = 1)
```

## Effects Plot for the Main Effects Model

Recall that the effects plot for the **interaction model** simply **shows a separate line for each combination of quantitative regressor and factor level because they interact and the effects cannot be separated**.

```{r}
# We have already seen this plot
# Compute effects
#ppgdpEffect <- Effect(c("ppgdp", "group"),
#                     mod = lmodi,
#                     xlevels = list(ppgdp = seq(1, 106000, len = #1000)))

# Plot the effects
plot(ppgdpEffect,
     rug=FALSE, grid=TRUE, multiline=TRUE,
     xlab = "log(ppgdp)",
     transform.x=list(ppgdp=list(trans=log, inverse=exp)),
     ticks.x =list(ppgdp = list(at= c(100, 1000, 5000, 30000))))
```
```{r}
#Another way to plot on log-scale
plot(ppgdpEffect, multiline=TRUE, grid=TRUE,rug=FALSE,
     axes=list(
       x=list(ppgdp=list(transform=list(trans=log, inverse=exp),
                        lab = "ppdgd, log-scale",
                        ticks=list(at=c(100,1000,5000,30000)),
                        lim = c(1,200000)
                         ))))
```

In the **main effects model**, the **effect of the factor level can be distinguished separately from the quantitative regressor**.

```{r}
# effect plot for log(ppgdp) by group
ppgdpEffect <- allEffects(mod = lmodm,
                     xlevels = list(ppgdp = seq(1, 106000, len = 1000)))

# effect plot of ppgdp on original scale for main effects model
plot(ppgdpEffect)
```

- The main effects model effects plot separates the effect of the factor (holding the regressor constant at its mean) and the effect of the regressor (which has only a single line since the effect is the same for all levels of the factor).

**Note:**

- The vertical bars/shading in each plot indicate +/- 1 standard error of the estimated effect size.
- The ppgdp effects plot is a weighted average of the fitted lines for the three levels of group.  
- The absolute scale doesn’t mean very much, but the shape is quite important.

