---
title: "Categorical Predictors Part 1"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---


```{r, setup, include=FALSE}
# This option means all R code is NOT evaluated
# Delete or comment out if you want to run 
# R commands when knitting
knitr::opts_chunk$set(eval = FALSE)
```

```{r, message = FALSE}
# Here we load libraries used in this file
# You may need to install if not already done so
library(alr4) #datasets from Weisberg text
```

# Categorical Predictors

Categorical predictors are often called factors.  

- Factors can have two levels (e.g., survived or died) or
- Multiple levels (single, married, divorced, widowed).

# Exploring Data for One-Factor Models: Example with United Nations Dataset

### 1. Load the `UN11` dataset below and read the help documentation to familiarize yourself with the data. How many observations are in the dataset? How many variables?

```{r}
data(UN11)
?UN11
```

We'll be working with following variables.

- **ppgdp**: the gross national product per person in U.S. dollars
- **fertility**: Average number of children per birth parent
- **lifeExpF**: Female life expectancy, years
- **pctUrban**: A percentage of population living in urban areas
- **group**: A factor with level **oecd** for countries that are members of the Organization for Economic Cooperation and Development, are located in **africa**, and **other** for all other countries.

If the only predictor is a factor, then the regression model is called a **one-factor** or **one-way** design.

```{r}
levels(UN11$group)
```

## Exploratory Data Analysis

### 2. Create side-by-side boxplots for the variable **lifeExpF** for each of the categories of the **group** variable and describe any interesting observations about the data.

```{r}
boxplot(??) 
```


## How do we include factor variables as regressors?

**Dummy** or **indicator variables** are used to include categorical predictors in a regression model.

- A dummy variable $U_j$ for factor level $j$ is 1 if an observation has level $j$, but $0$ otherwise.
- Note that -1 and 1, or 1 and 2, are sometimes used, but this is less common and more difficult to interpret.
- Assignment labels are mostly arbitrary and do not affect the results.

Since group has $d=3$ levels, the $j^{\mbox{th}}$ dummy variable $U_j$ for the factor $j=1,2, \ldots,d$ has $i^{\mbox{th}}$ value $u_{ij}$, for $i=1,2, \ldots,n$ given by 

$$u_{ij} = \left\{ \begin{array}{ll} 1 & \mbox{if } \mbox{group}_i=j^{\mbox{th}} \mbox{ category of group} \\
0 & \mbox{otherwise} \end{array} \right.$$

For the **group** variable, we set up three dummy variables

- $U_1$ (is country OECD?), 
- $U_2$ (is country Other?), and 
- $U_3$ (is country in Africa?).

### 3. Explain what the code below is doing? What happens if we remove the $+ 0$ from the commands?

```{r}
# with functions means: using this data frame, 
# perform the following actions ...
U1 <- with(UN11, (group == levels(group)[1]) + 0) 
U2 <- with(UN11, (group == levels(group)[2]) + 0)
U3 <- with(UN11, (group == levels(group)[3]) + 0)
head(data.frame(group = UN11$group, U1, U2, U3), 10)
```

## The one-factor model

Regression coefficients for factor variables in a one-way model are often called **effects**.

### 4. Let $Y$ denote the response variable `lifeExpF`. What would be problematic about using the linear model proposed below?

$$Y =  \beta_0 + \beta_1 U_1 + \beta_2 U_2 + \beta_3 U_3 + \epsilon$$

We only need $d-1$ dummy variables to represent $d$ levels.

- Any of the $d$ dummy variables could be excluded, though generally it is the first or last level.
- R drops the first level by default.
- The level dropped is known as the **reference** or **baseline level**.
- In R, this method for creating the dummy variables is known as the **treatment contrast**.


Other possibilities for avoiding linear dependence are: 

- Drop $\beta_0$ from the model.
- Assume $\sum_{i=1}^{d-1} \beta_i = 0$.  
  - This is known as a **sum contrast** and does NOT use dummy variables.
  - Interpretation is generally more difficult.

### 5. Using a treatment contrast with the first level of group as the reference level (oecd), what would be our model?



For the OECD group, since $U_2=U_3=0$, we have

$$\mbox{E}(\mbox{lifeExpF } \vert \mbox{ group=oecd})= \beta_0 + \beta_2 0+ \beta_3 0= \beta_0$$.


### 6. Find similar formulas for E(lifeExpF $\vert$ group=other) and E(lifeExpF $\vert$ group=africa).


## Fitting a One-Factor Model with R

R will fit a one-factor model if you: 

- Add the necessary dummy variables manually to the `formula` argument of `lm`.
- Include the factor variable in the formula of `lm` (dropping the dummy variable for the first level of the factor automatically).

```{r}
# First we add dummy variables to the dataframe
UN11$U1 = U1
UN11$U2 = U2
UN11$U3 = U3
```

```{r}
# fit overparameterized model
lm(lifeExpF ~ U1 + U2 + U3, data = UN11)
```

```{r}
# fit proper model
lm(lifeExpF ~ U2 + U3, data = UN11)
```

```{r}
# what R does by default
lm(lifeExpF ~ group, data = UN11)
```

### 7. Compare the output from the previous three models and commment on the differences, similarlities of each.

Using the second model above with dummy variables $U_2$ and $U_3$, we have the following:

$\widehat{\mbox{E}} (\mbox{lifeExpF }\vert\mbox{ group = oecd}) = \widehat{\beta}_0 = 82.45$, 

$\widehat{\mbox{E}} (\mbox{lifeExpF }\vert \mbox{ group = other}) = \widehat{\beta}_0 + \widehat{\beta}_2 = 75.33$

$\widehat{\mbox{E}} (\mbox{lifeExpF }\vert \mbox{ group = africa})= \widehat{\beta}_0 + \widehat{\beta}_3 = 59.79$. 

### 8. Interpret the practical meaning of the regression coefficents $\widehat{\beta}_0$,  $\widehat{\beta}_2$ and  $\widehat{\beta}_3$


### 9. Explain what the code below does, and compare the output to your previous answers.

```{r}
tapply(UN11$lifeExpF, UN11$group, mean)
```

## Visualizing the Model: Group Effect Plot

```{r}
lmod <- lm(lifeExpF ~ group, data = UN11)

# effects plot of one-way model
plot(allEffects(lmod, default.levels=50), ylim=c(55, 85),
     grid=TRUE, multiline=TRUE)
```