---
title: "Integration in R"
#date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

# Integrating using the `pracma` library

You will need to install the `pracma` package one time.
Then you will need to load it with `library(pracma)` each time you begin a new session in which you wish to use the `integral` and `integral2` functions.

```{r}
#install.packages("pracma")
library(pracma)
```

# Function of 1 variable

```{r}
f <- function(x) x^2 # first enter the function
integral(f, 0, 3) # enter x limits first and y limits last
```

# Function of 2 (or more) variables

```{r}
f <- function(x,y) x*y # first enter the function
integral2(f, 5, 30, 10, 30) # enter x limits first and y limits last
```

# Derivatives and Antiderivatives in R

You will need to run the first line of code one time to install the required `mosaicCalc` package

```{r}
#install.packages("mosaicCalc")
library(mosaicCalc)
der.sq <- D(x^2 ~ x)
anti.sq <- antiD(x^2 ~ x)
der.sq
anti.sq
```

# Example from Class

```{r}
int.x <- function(x) x*(6/5*x + 2/5) # first enter the function
mu.x <- integral(int.x, 0, 1) # enter x limits first and y limits last

int.y <- function(y) y*(6/5*y^2 + 3/5) # first enter the function
mu.y <- integral(int.y, 0, 1) # enter x limits first and y limits last

mu.x
mu.y
```

# E(XY)

```{r}
int.xy <- function(x,y) x*y*(6/5*(x + y^2)) # first enter the function
integral2(int.xy, 0, 1, 0, 1) # enter x limits first and y limits last
```
